---
title: "R Notebook"
output:
  word_document: default
  pdf_document: default
  html_notebook: default
---
1. Load the data into an appropriate data object of your choice (data frame or tibble). Note that the data file is compressed so you need to determine and appropriate way to uncompress the file as you read it (do not uncompress the file in your operating system). See the resources below to view an article that explains various ways to read compressed files.
```{r}
#Setting the directory
setwd("C:/Users/prith/OneDrive/Desktop/Assignment 4")
getwd()
#Unzipping the file
file <- unzip("bird_strikes.zip")
#Reading the file and placing it in 'data'
data <- read.csv(file, stringsAsFactors = F)
#data <-read.table(unz("bird_strikes.zip", "Bird Strikes.csv"), header=T, quote="\"", sep=",", fill =T, stringsAsFactors =F)
#Loading the tidyverse package
library(tidyverse)
#Checking top 10 rows
head(data, n=10)
```
2.  Write a function called strikesByYear() that returns a dataframe containing for each year the number of bird strike incidents.
```{r}
#Checking the type of FlightDate. Whether it is numeric or character 
#class(data1$FlightDate)
#data1 <- data
#data1$FlightDate <- as.Date(data1$FlightDate, "%m/%d/%Y")
#Loading the 'lubridate' function
library(lubridate)
#Creating a function named strikesByYear()
strikesByYear <- function(d){
#Converting the Flightdate in the format date(month, day and year)  
d$FlightDate <- as.Date(d$FlightDate, "%m/%d/%Y")
#Representing it by year, month and day
ymd(d$FlightDate)
#Taking only the year
d$FlightDate <- year(d$FlightDate)
#Using pipes we find the number of bird strikes in each year
#Selecting only the year and birdstrikes
x <- select(d, FlightDate) %>% 
#Grouping by FlightDate
group_by(FlightDate) %>%
#n() inside summarise is used to count and it is put in Bird_Strikes
summarise(Bird_Strikes = n())
#Returning the value
return (x)
} 

#Calling the function and passing Data in it
strikesByYear(data)
```

3. Write a function called leastStrikesInaYear() that returns the year which had the least bird strike incidents.
```{r}
#Creating a function named leastStrikesInaYear
leastStrikesInaYear <- function(data2){
#Converting the Flightdate in the format date(month, day and year)  
data2$FlightDate <- as.Date(data2$FlightDate, "%m/%d/%Y")
#Representing it by year, month and day
ymd(data2$FlightDate)
#Taking only the year
data2$FlightDate <- year(data2$FlightDate)
#Selecting the year and birdstrikes
x <- select(data2, FlightDate) %>%
#Using group_by to group them by the flightdate  
group_by(FlightDate) %>%
#n() inside summarise is used to count and it is put in Bird_Strikes
summarise(Bird_Strikes = n()) %>%
#Filer function is used to find the minimum value or the least value and it is saved in Bird_Strikes
filter(Bird_Strikes==min(Bird_Strikes))
#Returning the value
return (x)
}

#Calling the function and passing the data in it   
leastStrikesInaYear(data)

```

4. Write a function called strikesByType() that returns the number of birds strike incidents per type of wildlife (excluding unkown) in a dataframe. Store the return result in a dataframe called Strikes.

```{r}
#sum(is.na(data$Wildlife..Species)) - Checking if there are any NA values in the 'Wildlife..Species' column

#Creating a function named StrikesByType
strikesByType <- function(Strikes){
#str_detect(data$Wildlife..Species,"Unknown")
#Selecting the Wildlife..Species and BirdStrikes. Pipe operation is used  
x <- select(Strikes, Wildlife..Species) %>% 
#Using the group by statement and grouping by Wildlife..Species 
group_by(Wildlife..Species) %>%
#Using the filter statement and making sure no Unknown species are included and no NULL or "" species are included.  
filter(!(str_detect(Wildlife..Species,"Unknown") | Wildlife..Species=="")) %>% 
#n() inside summarise is used to count and it is put in Bird_Strikes
summarise(Bird_Strikes = n()) 
#Returning the value
return(x)
}
#Calling the function and passing the data in it
strikesByType(data)
```